Nengo: A Python tool for building large-scale functional brain models

### 2013-12-18 nengo.tex
Submission finalized, references fixed

Decoupling model creation and simulation
has been done previously
by PyNN \citep{davison2008}.
In PyNN, the same Python script
can be used to run a model
on four different simulators.
Nengo follows this programming model by
decoupling neural model creation and simulation,
which enables Nengo simulators
to allocate memory and schedule computations
in the most efficient ways possible.
Simulators are given a \texttt{Model}
as an argument;
this \texttt{Model} is a static symbolic description.
The simulator can take the model description
and build whatever data structures
are best suited to implement the simulation.

@@

The communication channel and Lorenz attractor
are small models that demonstrate
the principles of the NEF.
Their small size enables us to write
PyNN scripts that implement roughly
the same functionality
with Brian \citep{goodman2009}, NEURON \citep{hines2009},
and NEST \citep{eppler2008}.\footnote{We
  were unable to compile PCSim \citep{pecevski2009} on the
  machine on which we ran benchmarks.}
We ran each parameter set five times
on the same machine,
and plot the mean time elapsed
in Figure~\ref{fig:benchmarks}.
In most cases, the coefficient of variation
for the five sample times
is well below 0.1, except for two
outliers with coefficients of 0.18 and 0.22,
overall indicating that the reported means are robust.
The results, shown in Figure~\ref{fig:benchmarks}A and B,
suggest that all versions of Nengo are significantly
faster than the simulators accessible
through PyNN, especially
as the size of models increases.
This is likely due to Nengo's
use of factorized weight matrices,
rather than storing and computing with
the entire weight matrix
on each timestep.
While NEST and NEURON were not
run on multiple cores using message passing,
the reference simulator of Nengo
also only uses one CPU core.
The results further suggest that Nengo 2.0's
simulators are faster than Nengo 1.4's simulator.

@@

There are many other neural simulators
dedicated to building large-scale neural models
\citep{eppler2008,goodman2009,hines2009},
and many tools for simulating cognitive phenomena
with various levels of biologically plausibility
\citep{cooper1998,sun2001,anderson2004,franklin2007,
  aisa2008,dekamps2008,laird2012}.
However, Nengo is unique in being built
on a theoretical framework
that has enabled a cognitive architecture
(the Semantic Pointer Architecture)
that maintains a high level
of biological plausibility,
and has been validated
through the Spaun model and other past work.

The most closely related projects
in terms of software design
are PyNN \citep{davison2008}
and Topographica \citep{bednar2009},
both of which provide a high-level scripting
interface to low-level neural simulators.
PyNN in particular has a similar to the
high-level object model in Nengo,
and provides a convenient interface
to the three most widely used
neural simulators, according
to a survey by \citet{hanke2011}.

>>>>

Decoupling model creation and simulation
has been done previously
by PyNN \citep{davison2009}.
In PyNN, the same Python script
can be used to run a model
on four different simulators.
Nengo follows this programming model by
decoupling neural model creation and simulation,
which enables Nengo simulators
to allocate memory and schedule computations
in the most efficient ways possible.
Simulators are given a \texttt{Model}
as an argument;
this \texttt{Model} is a static symbolic description.
The simulator can take the model description
and build whatever data structures
are best suited to implement the simulation.

@@

The communication channel and Lorenz attractor
are small models that demonstrate
the principles of the NEF.
Their small size enables us to write
PyNN scripts that implement roughly
the same functionality
with Brian \citep{goodman2008}, NEURON \citep{hines2009},
and NEST \citep{eppler2009}.\footnote{We
  were unable to compile PCSim \citep{pecevski2009} on the
  machine on which we ran benchmarks.}
We ran each parameter set five times
on the same machine,
and plot the mean time elapsed
in Figure~\ref{fig:benchmarks}.
In most cases, the coefficient of variation
for the five sample times
is well below 0.1, except for two
outliers with coefficients of 0.18 and 0.22,
overall indicating that the reported means are robust.
The results, shown in Figure~\ref{fig:benchmarks}A and B,
suggest that all versions of Nengo are significantly
faster than the simulators accessible
through PyNN, especially
as the size of models increases.
This is likely due to Nengo's
use of factorized weight matrices,
rather than storing and computing with
the entire weight matrix
on each timestep.
While NEST and NEURON were not
run on multiple cores using message passing,
the reference simulator of Nengo
also only uses one CPU core.
The results further suggest that Nengo 2.0's
simulators are faster than Nengo 1.4's simulator.

@@

There are many other neural simulators
dedicated to building large-scale neural models
\citep{eppler2009,goodman2008,hines2009},
and many tools for simulating cognitive phenomena
with various levels of biologically plausibility
\citep{cooper1998,sun2001,anderson2004,franklin2007,
  aisa2008,dekamps2008,laird2012}.
However, Nengo is unique in being built
on a theoretical framework
that has enabled a cognitive architecture
(the Semantic Pointer Architecture)
that maintains a high level
of biological plausibility,
and has been validated
through the Spaun model and other past work.

The most closely related projects
in terms of software design
are PyNN \citep{davison2009}
and Topographica \citep{bednar2009},
both of which provide a high-level scripting
interface to low-level neural simulators.
PyNN in particular has a similar to the
high-level object model in Nengo,
and provides a convenient interface
to the three most widely used
neural simulators, according
to a survey by \citet{hanke2011}.
